---
title: "Tutorial"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
# Need up to date packrat package to get it to install dependencies properly 
# devtools::install_github("rstudio/packrat")
library(learnr)
library(janitor)
library(tidyverse)
knitr::opts_chunk$set(echo = FALSE)
spending_data <- read_rds("csc_data_spend_imd.rds") %>%
  pivot_wider(names_from = description, values_from = value) %>%
  clean_names() %>%
  dplyr::rename(cla_exp = 4, sg_exp = 5, ehfs_exp = 6, pop = 7, imd = 8) %>%
  mutate(
    cla_exp = (cla_exp*100000) / pop
  ) %>%
  filter(year == 2015)
  

```


## Starting out with ggplot2

### A basic correlation plot

Let's start by loading in the data and telling R that we want to build a scatterplot, using the `ggplot()` initialisation function, which tells R we're building a ggplot graph, followed by a geometric object for plotting points, `geom_point()`. We have some data already loaded about children's services spending and deprivation.

Let's start by visualising the correlation between Indices of Multiple Deprivation score, named `imd` in our data, and safeguarding spending per child, named `sg_exp` in our data. Click 'run' to run the code, then move to the next exercise.


```{r ggplot-1, exercise=TRUE}

ggplot(spending_data) +
  geom_point(aes(x = imd, y = sg_exp))

```

## Creating your own correlation plot

Now try creating your own scatterplot by replacing the `___` in the code with the correct data and variable names. Instead of `sg_exp`, this time we'll look at early help and family support spend per child. This is named `ehfs_exp` in the data. As a reminder, this is the code we ran in the last exercise.

```{r, eval=FALSE, echo=TRUE}

ggplot(spending_data) +
  geom_point(aes(x = imd, y = sg_exp))

```

*Change the `___`s to the correct variable names to get the code to create the scatterplot*

```{r ggplot-2, exercise=TRUE}

ggplot(___) +
  geom_point(aes(x = ___, y = ___))

```

```{r ggplot-2-solution}

ggplot(spending_data) +
  geom_point(aes(x = imd, y = ehfs_exp))

```

## Adding a static aesthetic to the plot

Let's say that we want our scatterpoints to be a bit larger than they are. We can do that by adding a static aesthetic argument to the plot. These arguments come after the `aes()` argument in the `geom_*()` function. You can find the list of aesthetic arguments that each geometric object can take, including which ones are required, by typing `?geom_*` into the console (replace * with the name of the geometric object function). `geom_point` understands the following: `alpha`, `colour`, `fill`, `group`, `shape`, `size`, and `stroke`. Let's use `size` to make the size of our points larger. 

*Try replacing the `size` argument with different values to see what you think looks best.*

```{r ggplot-3, exercise=TRUE}

ggplot(spending_data) +
  geom_point(aes(x = imd, y = ehfs_exp), size = 4)

```

## Adding a dynamic aesthetic to the plot

If we put an aesthetic argument within the `aes()` parentheses `ggplot2` expects us to link it to a variable that changes. We did this with the `y` and the `x` aesthetics, which specify the data that determines where on the Y and X axes each point will be plotted. It can be useful to use things like size (`size`), colour (`colour`), and transparency (`alpha`) to include more information in a plot or to make it more aesthetically appealing. 

Let's add a dynamic colour aesthetic to our plot that is linked to local authorities' IMD score by adding `colour = imd` within the `aes()` brackets.

*Try moving the `colour = imd` argument outside the `aes()` brackets and see what changes.*

```{r ggplot-4, exercise=TRUE}

ggplot(spending_data) +
  geom_point(aes(x = imd, y = ehfs_exp, colour = imd), size = 4)

```

## Changing themes and scales

We can also add additional functions to our ggplot code to change its appearance. For example, in the below code, the colour scale for IMD score has been changed from black-to-blue to blue to red using the `scale_colour_gradient()` function. We can also add labels using the `ylab()` and `xlab()` functions.

*Try changing the labels and the colour choices on the new colour scale.*

```{r ggplot-5, exercise=TRUE}

ggplot(spending_data) +
  geom_point(aes(x = imd, y = ehfs_exp, colour = imd), size = 4) +
  scale_color_gradient(low = "blue", high = "red") +
  ylab("Early Help and Family Support Expenditure") +
  xlab("Indices of Multiple Deprivation Score")

```